Deep Neural Architectures for Discourse Segmentation in E-Mail Based Behavioral Interventions

### 2018-11-15 paper-amia-summit-2019/manuscript.tex
update area under precision-recall values and update manuscript

The goal of this study is to assess the effectiveness of deep learning methods for the task of automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of qualitative analysis of this type of clinical communications. Specifically, for this study, we utilized the data from MENU GenY (Making Effective Nutrition Choices for Generation Y) \cite{alexander2017motivations}, a recent technology-based public health intervention that relies on personalized e-Coaching to encourage increased fruit and vegetable intake among young adults, aged 21--30. The goal of MENU GenY was to develop a better coding scheme for e-Coaching communications aimed at improving GenY eating habits. Segmentation of clinical conversation in the context of electronically delivered interventions into groups of MI behaviors, is traditionally performed manually by MI researchers, which significantly slows down their qualitative analysis. This paper is the first work to evaluate the empirical effectiveness of deep learning architectures in addressing \textit{the problem of discourse segmentation in the context of e-mail based behavioral interventions}.

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include a total of 3,138 segmented and annotated MI behaviors \hl{shown in Table}~\ref{tab:datastat}. Each session represents an MI intervention delivered via email. We consider segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a label of 1 or 0 to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances. In this study, we experimented with traditional machine learning methods, such as Conditional Random Field (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing instances were created based on a sliding window of $2n$ words or punctuation marks over each position in a given input sequence, such that each instance consists of a pair of the next $n$ words or punctuation marks and prior $n$ words or punctuation marks, including the current word or punctuation mark. Each sample is classified into either a ``new segment'' or a ``same segment'' based on whether there should be a segment break after the current word or not. In the case of CRF, BRNN and CRNN models, an email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output of a model. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1.

@@

\subsection*{\textit{Features}}

We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings or lexical features, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used POS features for our experiment.\cite{liu2005using,treviso2017sentence} To extract POS features, we pre-processed e-Coaching emails using the NLTK POS tagger. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, were also used as a feature, since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text, in which individual words are as the basic units of semantics. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embedding is one such representation, when each word is associated with a real-valued vector in a high dimensional vector space. Distributed representations of words have been shown to capture semantic, syntactic and morphological properties of words.\cite{pennington2014glove, mikolov2013distributed} For experiments reported in this paper, we utilized embeddings estimated on Google News corpus consisting of 1.6 billion words pre-trained using word2vec word embedding method.\footnote{\label{fn:word2vec}https://code.google.com/p/word2vec/} When words or punctuation marks are not found in the pre-trained word vectors, we utilized word embeddings trained with our e-Coaching email corpus, which we refer to as corpus-based embeddings. CRF utilized lexical features, which correspond to words in e-Coaching emails.

@@

where $Z_x$ is a normalization factor, $f_k(y_{t-1}, y_t, x, t)$ is a feature function, and $\lambda_k$ is a learned weight associated with feature $f_k$. The optimal output sequence $y^*$ for an input sequence $x$, $y^* = {arg\,max}_y p(y|x)$, is obtained efficiently using the Viterbi algorithm. In our experiments, the following features were utilized in conjunction with CRF models: i) current word or punctuation ii) next and previous 3 words or punctuations iii) binary feature indicating whether a word or punctuation is a special character (';', '?', '.', ',', '!', ':', etc.) or not iv) binary feature indicating whether a word is a title word or not (e.g. ``The'' is a title word but ``the'' is not) v) POS tags

@@

\subsection*{\textit{Evaluation metrics}}
We report standard metrics of precision, recall and F1-measure to evaluate the performance of the classifiers.\cite{aas1999text} Accuracy is not reported as a performance metric because it is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem, when highly unbalanced datasets are involved. The results are reported based on 5 fold cross-validation and weighted macro-averaging over the folds. We also estimate the area under the receiving operating characteristics (ROC) curve\cite{kumar2011receiver} (AUC) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets. \cite{hu2015kernelized}

@@

As follows from Table~\ref{tab:result_base}, CRNN outperforms all other models in terms of recall and F1-measure achieving 0.797 recall with 0.785 F1-measure for new segment detection. CRNN also shows superior performance in all performance metrics for weighted average over ``new segment'' and ``same segment'' classes. BRNN demonstrates the lowest performance among all models in terms of precision and F1-Measure. On the other hand, MLP has the highest precision of 0.836 when word embeddings or lexical features are used to identify ``new segment''. CRF achieves 0.733 F1-Measure in new segment class and 0.984 F1-Measure overall, which corresponds to the second highest performance in identifying ``new segment'' as well as a weighted average over both classes. Experimental results indicate that performance of all classifiers as a weighted average over both classes is remarkably higher compared to ``new segment'' class, which is expected since 96.95\% of instances belong to the ``same segment'' class and 99.3\% of them are correctly classified. For example, as a weighted average over ``new segment'' and ``same segment'' classes, CRNN achieves 27.23\%, 23.71\% and 25.61\% higher precision, recall and F1-measure, respectively, compared to the ``new segment'' detection.

@@

@@

Table~\ref{tab:result_roc} shows the impact of individual features as well as their combination on the segmentation task. Although CRF provides the second highest classification results, it shows best AUC values, achieving AUC 0.966 when only lexical features are used and AUC 0.994 when a combination of lexical, punctuation and POS features are used. Influence of the punctuation and POS features is also consistent in AUC values, which increase by 2.9\%, 4.31\%, 6.63\% and 2.18\% for CRF, MLP, BRNN and CRNN methods, respectively, when all features are used together. Individually, POS and punctuation features also improve the performance of all classifiers except CRNN when word embeddings are used with POS features. CRF and MLP achieve their highest AUCs when all features are used together. On the other hand, BRNN and CRNN demonstrate the highest AUCs when word embeddings are combined with punctuation features. \\

\begin{table}[ht]
\centering
\caption{\textbf{AUC values of all classifiers demonstrating the impact of word embeddings, punctuation and POS features on e-Coaching email segmentation. Highest AUC value for each feature set across all models is highlighted in boldface.}}
\label{tab:result_roc}
 \begin{tabular}{|l|l|l|l|l|}
  \hline
\textbf{Features} & \textbf{CRF} & \textbf{MLP}  & \textbf{BRNN} & \textbf{CRNN} \\ \hline      
 word embeddings only & \textbf{0.966} & 0.951 & 0.920 & 0.965 \\ \hline
 word embeddings + POS & \textbf{0.973} & 0.953 & 0.920 & 0.962 \\ \hline
 word embeddings + punctuation & \textbf{0.993} & 0.989 & 0.982 & 0.988 \\ \hline
 all features & \textbf{0.994} & 0.992 & 0.981 & 0.986 \\ \hline
  \end{tabular}
\end{table}

@@

Punctuation mark and POS features resulted in significant improvement in the performance of machine learning and all deep learning methods. Especially, punctuation features have the higher individual impact on model performance compared to POS features. In almost all cases, every method performs better, when word embeddings are used in conjunction with punctuation and POS features. This results also indicate that segmentation performance might be improved by adding additional relevant features.

>>>>

The goal of this study is to assess the effectiveness of deep learning methods for the task of automated segmentation of e-Coaching emails into textual fragments corresponding to individual behaviors, which is the first step of qualitative analysis of this type of clinical communications. Specifically, for this study, we utilized the data from MENU GenY (Making Effective Nutrition Choices for Generation Y) \cite{alexander2017motivations}, a recent \hl{email-based} public health intervention that relies on personalized e-Coaching to encourage increased fruit and vegetable intake among young adults, aged 21--30. The goal of MENU GenY was to develop a better coding scheme for e-Coaching communications aimed at improving GenY eating habits. Segmentation of clinical conversation in the context of electronically delivered interventions into groups of MI behaviors, is traditionally performed manually by MI researchers, which significantly slows down their qualitative analysis. This paper is the first work to evaluate the empirical effectiveness of deep learning architectures in addressing \textit{the problem of discourse segmentation in the context of e-mail based behavioral interventions}.

@@

\section*{Methods}
\subsection*{\textit{Data collection}}
The experimental dataset for this work was constructed from 49 e-Coaching sessions, which include \hl{330 and 281 emails from ecoaches and patients, respectively }(Table~\ref{tab:datastat}). \hl{Each session represents an MI intervention delivered via email. Emails were segmented into 3,138 text fragments and annotated using 115 MYSCOPE}~\cite{carcone2013provider} \hl{behavior codes. We consider email segmentation as a binary classification problem, in which each word or punctuation mark is annotated with a label of 1 or 0 to indicate whether it precedes a new segment or not. In total, the dataset consists of 95,777 words and 7,140 punctuation marks and includes 3,138 ``new segment'' and 99,779 ``same segment'' instances.} In this study, we experimented with traditional machine learning methods, such as Conditional Random Field (CRF)\cite{lafferty2001conditional} and deep learning methods, such as Multi-Layer Perceptron (MLP),\cite{rumelhart1986learning} Bidirectional Recurrent Neural Network (bidirectional RNN or BRNN)\cite{schuster1997bidirectional} and Convolutional Recurrent Neural Network (CRNN).\cite{treviso2017sentence} For an MLP model, training and testing instances were created based on a sliding window of $2n$ words or punctuation marks over each position in a given input sequence, such that each instance consists of a pair of the next $n$ words or punctuation marks and prior $n$ words or punctuation marks, including the current word or punctuation mark. Each sample is classified into either a ``new segment'' or a ``same segment'' based on whether there should be a segment break after the current word or not. In the case of CRF, BRNN and CRNN models, an email was taken as an input sequence, POS tags and word embeddings of each word or punctuation mark were used as input and binary labels (1 or 0) corresponding to ``new segment'' and ``same segment'' classification decisions were considered as the model output of a model. In the gold standard, words or punctuations within the same segment were assigned the label of 0 and the last word or punctuation mark of a segment were assigned the label of 1. \\

@@

\subsection*{\textit{Features}}
We utilized three types of features in conjunction with CRF, MLP, BRNN and CRNN methods: word embeddings or lexical features, punctuation and POS features. Since POS tags have been shown to be effective semantic abstractions of individual words, we used POS features for our experiment.\cite{liu2005using,treviso2017sentence} To extract POS features, we pre-processed e-Coaching emails using the NLTK POS tagger. Punctuation marks, which correspond to one of the symbols \{`.', `,', `!', `?', `:', `;'\} between a pair of words, were also used as a feature, since punctuation marks designate the boundary of a sentence, clause and phrase and often also correspond to a segment boundary.\cite{cho2002text} For natural language processing (NLP) tasks, inputs are received as a text, in which individual words are as the basic units of semantics. Therefore, it is important to represent a word in such a way that it carries all relevant information. Word embedding is one such representation, when each word is associated with a real-valued vector in a high dimensional vector space. Distributed representations of words have been shown to capture semantic, syntactic and morphological properties of words.\cite{pennington2014glove, mikolov2013distributed} For experiments reported in this paper, we utilized embeddings estimated on Google News corpus consisting of 1.6 billion words pre-trained using word2vec word embedding method.\footnote{\label{fn:word2vec}https://code.google.com/p/word2vec/} When words or punctuation marks are not found in the pre-trained word vectors, we utilized word embeddings trained with our e-Coaching email corpus, which we refer to as corpus-based embeddings. CRF utilized lexical features, which correspond to words in e-Coaching emails.

@@

where $Z_x$ is a normalization factor, $f_k(y_{t-1}, y_t, x, t)$ is a feature function, and $\lambda_k$ is a learned weight associated with feature $f_k$. The optimal output sequence $y^*$ for an input sequence $x$, $y^* = {arg\,max}_y p(y|x)$, is obtained efficiently using the Viterbi algorithm. In our experiments, the following features were utilized in conjunction with CRF models: i) current word or punctuation ii) next and previous 3 words or punctuations iii) binary feature indicating whether a word or punctuation is a special character (';', '?', '.', ',', '!', ':', etc.) or not iv) binary feature indicating whether a word is a title word or not (e.g. ``The'' is a title word but ``the'' is not) v) POS tags.

@@

\subsection*{\textit{Evaluation metrics}}
We report standard metrics of precision, recall and F1-measure to evaluate the performance of the classifiers.\cite{aas1999text} Accuracy is not reported as a performance metric because it is highly sensitive to the prior class probabilities and does not fully describe the actual difficulty of the decision problem, when highly unbalanced datasets are involved. The results are reported based on 5 fold cross-validation and weighted macro-averaging over the folds. We also estimate the area under \hl{the precision-recall curve (AUPR) metric due to its effectiveness in measuring the quality of binary classifiers for imbalanced datasets.} \cite{davis2006relationship}

@@

@@

As follows from Table~\ref{tab:result_base}, CRNN outperforms all other models in terms of recall and F1-measure achieving 0.797 recall with 0.785 F1-measure for new segment detection. CRNN also shows superior performance in all performance metrics for weighted average over ``new segment'' and ``same segment'' classes. BRNN demonstrates the lowest performance among all models in terms of precision and F1-Measure. On the other hand, MLP has the highest precision of 0.836 when word embeddings or lexical features are used to identify ``new segment''. CRF achieves 0.733 F1-Measure in new segment class and 0.984 F1-Measure overall, which corresponds to the second highest performance in identifying ``new segment'' as well as a weighted average over both classes. Experimental results indicate that performance of all classifiers as a weighted average over both classes is remarkably higher compared to ``new segment'' class, which is expected since 96.95\% of instances belong to the ``same segment'' class and 99.3\% of them are correctly classified. For example, as a weighted average over ``new segment'' and ``same segment'' classes, CRNN achieves 27.23\%, 23.71\% and 25.61\% higher precision, recall and F1-measure, respectively, compared to the ``new segment'' detection. \\

@@

\begin{table}[ht]
\centering
\caption{\textbf{\hl{Area under the precision-recall curve (AUPR) values of all classifiers demonstrating the impact of word embeddings, punctuation and POS features on e-Coaching email segmentation. Highest AUPR value for each feature set across all models is highlighted in boldface.}}}
\label{tab:result_aupr}
 \begin{tabular}{|l|l|l|l|l|}
  \hline
\textbf{Features} & \textbf{CRF} & \textbf{MLP}  & \textbf{BRNN} & \textbf{CRNN} \\ \hline      
 word embeddings only & 0.780 & 0.736 & 0.655 & \textbf{0.818} \\ \hline
 word embeddings + POS & 0.797 & 0.746 & 0.647 & \textbf{0.798} \\ \hline
 word embeddings + punctuation & \textbf{0.876} & 0.835 & 0.774 & 0.874 \\ \hline
 all features & \textbf{0.877} & 0.842 & 0.770 & 0.867 \\ \hline
  \end{tabular}
\end{table}

Table~\ref{tab:result_aupr} shows the impact of individual features as well as their combination on the segmentation task. \hl{Influence of the punctuation and POS features is also consistent in AUPR values, which increase by 12.44\%, 14.4\%, 17.56\% and 5.99\% for CRF, MLP, BRNN and CRNN methods, respectively, when all features are used together. Individually, POS and punctuation features also improve the performance of all classifiers except BRNN and CRNN when POS features are used. CRF achieved the highest AUPR when all features are used together. On the other hand, CRNN demonstrates the highest AUPR when word embeddings are used individually as well as combined with POS features.}

@@

Punctuation mark and POS features resulted in significant improvement in the performance of machine learning and all deep learning methods. Especially, punctuation features have the higher individual impact on model performance compared to POS features. \hl{In all cases, CRF and MLP methods performed better, when word embeddings are used in conjunction with punctuation and POS features. Punctuation features improved the performance of BRNN and CRNN while POS features lowered their AUPR values. We believe that the BRNN and CRNN performed poorly with POS features because POS tagging is a supervised learning solution that uses features like the previous word and next word. Since we already considered neighbor words by utilizing bi-directional RNN, it failed to achieve good results with redundant information. We also observed that MLP achieved highest precision which may be related to the fact that MLP poorly learned new segment class and misclassiified segment words as non-segment words.}
